NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-11-21 Manuscript/manuscriptPS.tex
Update text

    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.45 & 0 & 110:22 & 4926  & 40 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298KwithK}  \\
    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.89 & 0 & 110:22 & 4946  & 79 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298KwithK}  \\

@@

    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.44 & 0 & 110:22 & 4965  & 39 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298KwithNa}  \\
    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.89 & 0 & 110:22 & 4932  & 79 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298KwithNa}  \\

@@

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/structures.eps}
  \caption{\label{HGstructuresPSandPC}
    Overlayed snapshots from glycerol backbone and headgroup region from 
    different simulations of PS lipids.
  }
\end{figure}

@@

The headgroup and glycerol backbone of PS lipids show wide variety between different simulation
models in the order parameters and structures (Figs. \ref{HGorderParametersPS} and \ref{HGandGLYstructuresPS}),
as previously observed also for PC lipids \cite{botan15}.
The models perform generally less well for PS lipids than for PC lipids in the previous study
(Figs. 2 and 4 in Ref.~\cite{botan15} vs. Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS}).
Therefore, interpretation of structural differences between PC and PS headgroups
from simulations is not straightforward.

The best performing models,
Slipids, CHARMM36 and CHARMM36ua, reproduce the larger forking of the $\alpha$-carbon 
and the Slipids model reproduces also the lower of the $\beta$-carbon order parameter
when comparing the PS results to PC (Fig. 2 in Ref.~\citenum{botan15} vs. Fig. \ref{HGorderParametersPS}).
Interestingly, the dihedral angle distributions of C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ show
a single narrow maximum close to 120$^{\circ}$ in the best three models, while other models give several maxima
in different locations (Fig. \ref{dihedralsHG}). The restricted motion is also
visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} a) and \ref{HGandGLYstructuresPS})
suggesting that the rotation of carboxyl group is limited in the serine headgroup.
In addition, the comparison of N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral between PC and PS headgroups
from CHARMM36 simulations, which are in best agreement with experiments for both lipids,
shows a more asymmetric and restricted
rotations around this bond (Figs.~\ref{HGstructuresPSandPC} b) and \ref{dihedralsHGpc}).
These results might manifest the increased rigidity anticipated from the early experimental studies \cite{browning80,buldt81}.
Also the dihedral distributions and sampled conformations of glycerol backbone region
from different models are shown in the supplementary information
(Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),
but further analysis is beyond the scope of this work which the PS headgroup.

%which can be related to the differences in the dihedral angle distributions 
%of C1-C2-C3-O31 and C2-C3-O31-C31 (Fig.~\ref{dihedralsGLY}).

%  One possible conclusion could be the following:
%  The main differences between the models in the headgroup region are observed
%  for dihedrals C12-C11-O12-P and C11-C12-C13-O1A.
%  CHARMM36, CHARMM36UA and Slipids give very similar results to the
%  dihedral C11-C12-C13-O1A, which is close to the $\beta$-carbon.
%  The order parameters of $\beta$-carbons for these three models
%  are in best argeement with the experiments in figure \ref{HGorderParametersPS}.
%  On the other hand, Gromos-CKP models give better order parameters
%  for $\alpha$-carbon than Slipids, CHARMM36 or CHARMM36UA. 
%  In conclusion, the suggestion would be that the single peak for
%  observed at 120 degrees in CHARMMs and Slipids would
%  be more realistic for C11-C12-C13-O1A dihedral, while the single peak at 180 degrees observed
%  in CKP models and in Berger would be most realistic for C12-C11-O12-P dihedral.
%}

@@

\subsection{Counterion binding and interactions between PC and PS headgroups}\label{ciBINDINGsection}
Membranes containing PS lipids are always accomppanied with counterions which
modulate electrostatic interactions between lipids and other biomolecules. 
Counterions are also suggested to screen the repulsion between charged lipid headgroups 
in MD simulations and reduce the area per lipid of PS bilayers to be smaller than in PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06}. 
The counterion density profiles along membrane normal 
show significant differences between simulation models in both binding affinity
and distribution of ions in the interface (Fig. \ref{NAdensPOPS}).
The experimental area per lipid (62.7 $\AA^2$) \cite{pan14} 
is reproduced only in Gromos-CKP simulations and in the MacRog simulation
with potassium counterions, while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
The counterion binding and concomintant electrostatic screening of the headgroup
repulsion does not fully explain the low area per molecule values
because the MacRog simulation with strongest sodium binding
(the lowest concentrations in bulk water) gives the same area per molecule
as CHARMM36ua simulation with significantly weaker counterion binding
affinity.
%CHARMM36, CHARMM36ua and Slipid models give
%smaller area per lipid than Gromos-CKP models with similar counterion binding affinity.
On the other hand, changing counterions from sodium to potassium, having weaker binding
affinity, increases the area per molecule from 53 $\AA^2$ to 63 $\AA^2$ in MacRog simulations.
%with potassium counterions and weaker
%binding affinity give larger area per molecule 
%increasing counterion binding affinity reduces the area per molecule
%when comparing between sodium and potassium in 
%or between
%ff99 and jc ions in Lipid 17 simulations.
In conclusion, the results are in line with the previous study suggesting that the
low area per molecule in PS lipid bilayers originate from the combination
of both counterion binding and hydrogen bonding network between lipid
headgroups \cite{petrache04}.
%, Berger and Lipid17/JC simulations.
%Slipids, Berger, MacRog, and Lipid17 exhibit a clear single maxima in the
%counterion density, but slightly at different distances from the bilayer
%center. Other models exhibit more complicated distribution, often including
%two local maxima in the density profile.
%CHARMM36, CHARMM36ua and Gromos-CKP models exhibit two local maxima in counterion
%density, while a single maxima is observed in the other models. 
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
  \caption{\label{NAdensPOPS}
    Counterion densities of POPS lipid bilayer along the membrane normal from
    simulations with different force fields.
  }
\end{figure}

@@

\begin{figure*}[]
  \centering
  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
  \caption{\label{changesWITHCaClPS}
    Changes of POPC (left) and POPS (right) headgroup order parameters from POPC:POPS (5:1) mixture
    as a function CaCl$_2$ concentration from experiments \citenum{roux90} and different simulations
    at 298K (except the data for Berger model is from simulation of POPC:POPS (4:1) mixture at 310K \cite{ollila07a,melcrova16}). 
    The order parameter values from systems without calcium are set as the zero point of y-axis,
    except for the $\alpha$-carbon order parameter of POPS (bottom, right) for which the both order parameters are shifted
    such that the lower order parameter is zero without additional ions to correctly illustrate
    the forking with different concentrations of calcium.
    Potassium counterions are used in MacRog simulations and sodium counterions in Lipid14/17 simulations.
    In CHARMM36 and Berger simulation with added calcium, the charge is neutralized with calcium and monovalent counterions are not present.
  }
  \todo{Upcoming simulations with original CHARMM36 have been mentioned in the blog:
  http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1520090718976\#c5569269391707740056} \\
\end{figure*}

\begin{figure}[]
  \centering
  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixture.eps}
  \caption{\label{CAdensPCPSmixture}
    Ca2+ density profiles from simulations.
  }
  \todo{The CHARMM results are mass densities, number densities should be used when the data by Jesper Madsen is available.} \\
  \todo{Should we include also counterions into the plot?} \\
  \todo{Figure needs general improvement.}
\end{figure}

>>>>

    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.45 & 0 & 110:22 & 4926  & 40 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298Kwith450mMK}  \\
    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.89 & 0 & 110:22 & 4946  & 79 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298Kwith890mMK}  \\

@@

    POPC:POPS (5:1)/Na$^+$  & CHARMM36 \cite{klauda10,venable13} &0.44 & 0 & 110:22 & 4965  & 39 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298Kwith440mMNa}  \\
    POPC:POPS (5:1)/Na$^+$  & CHARMM36 \cite{klauda10,venable13} &0.89 & 0 & 110:22 & 4932  & 79 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298Kwith890mMNa}  \\

@@

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/structures.eps}
  \caption{\label{HGstructuresPSandPC}
    Overlayed snapshots from CHARMM36 simulations in best agreement with experiments
    to demonstrate the conformational fluctuations around
    a) C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ and  O$_\alpha$-C$_\alpha$-C$_\beta$-N
    of PS headgroup and b) N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
    dihedrals of PS and PC headgroups.
  }
  \todo{We need atom labeling for b). Similar to the one in a) would be good} \\
\end{figure}

@@

The headgroup and glycerol backbone of PS lipids show wide variety between different simulation
models in the order parameters and structures (Figs. \ref{HGorderParametersPS} and \ref{HGandGLYstructuresPS}),
as previously observed also for PC lipids \cite{botan15}.
The models perform generally less well for PS lipids than for PC lipids in the previous study
(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs. 2 and 4 in Ref.~\cite{botan15}).
Therefore, the interpretation of structural differences between PC and PS headgroups
from simulations is not straightforward.

The best performing models,
Slipids, CHARMM36 and CHARMM36ua, reproduce the larger forking of the $\alpha$-carbon 
and the Slipids model reproduces also the lower of the $\beta$-carbon order parameter
when comparing the PS results to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}).
Interestingly, the dihedral angle distributions of C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ show
a single narrow maximum close to 120$^{\circ}$ in the best three models, while other models give several maxima
in different locations (Fig. \ref{dihedralsHG}). The restricted motion is also
visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} a) and \ref{HGandGLYstructuresPS})
suggesting that the rotation of carboxyl group is limited in the serine headgroup.
In addition, the CHARMM36 simulations, in best agreement with experiments for both lipids,
show more asymmetric and restricted N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral distribution 
in PS headgroup than in PC (Figs.~\ref{HGstructuresPSandPC} b) and \ref{dihedralsHGpc}).
These results might manifest the increased rigidity anticipated in the early experimental studies \cite{browning80,buldt81}.
Also the dihedral distributions and sampled conformations of glycerol backbone region
significantly vary between different simulations models 
(Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),
but further analysis is beyond the scope of this work which the PS headgroup.

The observed conformations of PS headgroup may be useful when interpreting experiments and
can guide the further force field development. However, more accurate simulations are
required to confirm the results because the simulated PS headgroup order parameters
are not within the experimental error bars in any of the tested models.

@@

\subsection{Counterion binding and interactions between PC and PS headgroups}\label{ciBINDINGsection}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
  \caption{\label{NAdensPOPS}
    Counterion densities of POPS lipid bilayer along the membrane normal from
    simulations with different force fields.
  }
\end{figure}

\begin{figure*}[ht]
  \centering
  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
  \caption{\label{changesWITHCaClPS}
    Changes of POPC (left) and POPS (right) headgroup order parameters from POPC:POPS (5:1) mixture
    as a function CaCl$_2$ concentration from experiments \citenum{roux90} and different simulations
    at 298K (except the data for Berger model is from simulation of POPC:POPS (4:1) mixture at 310K \cite{ollila07a,melcrova16}). 
    The order parameter values from systems without calcium are set as the zero point of y-axis,
    except for the $\alpha$-carbon order parameter of POPS (bottom, right) for which the both order parameters are shifted
    such that the lower order parameter is zero without additional ions to correctly illustrate
    the forking with different concentrations of calcium.
    Potassium counterions are used in MacRog simulations and sodium counterions in Lipid14/17 simulations.
    In CHARMM36 and Berger simulation with added calcium, the charge is neutralized with calcium and monovalent counterions are not present.
  }
  \todo{Upcoming simulations with original CHARMM36 have been mentioned in the blog:
  http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1520090718976\#c5569269391707740056} \\
\end{figure*}

\begin{figure}[ht]
  \centering
  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixture.eps}
  \caption{\label{CAdensPCPSmixture}
    Ca2+ density profiles from simulations.
  }
  \todo{The CHARMM results are mass densities, number densities should be used when the data by Jesper Madsen is available.} \\
  \todo{Should we include also counterions into the plot?} \\
  \todo{Figure needs general improvement.}
\end{figure}

Membranes containing PS lipids are always accomppanied with counterions which
modulate electrostatic interactions between lipids and other biomolecules. 
Counterions are also suggested to screen the repulsion between charged lipid headgroups 
in MD simulations and reduce the area per lipid of PS bilayers to be smaller than in PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06}. 
The counterion density profiles along membrane normal 
show significant differences between simulation models in both binding affinity
and distribution of ions in the interface (Fig. \ref{NAdensPOPS}).
The experimental area per lipid (62.7 $\AA^2$) \cite{pan14} 
is reproduced only in Gromos-CKP simulations and in the MacRog simulation
with potassium counterions, while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
The counterion binding and concomintant electrostatic screening of the headgroup
repulsion does not fully explain the low area per molecule values
because the MacRog simulation with strongest sodium binding
(the lowest concentrations in bulk water) gives the same area per molecule
as CHARMM36ua simulation with significantly weaker counterion binding
affinity.
%CHARMM36, CHARMM36ua and Slipid models give
%smaller area per lipid than Gromos-CKP models with similar counterion binding affinity.
On the other hand, changing counterions from sodium to potassium, having weaker binding
affinity, increases the area per molecule from 53 $\AA^2$ to 63 $\AA^2$ in MacRog simulations.
%with potassium counterions and weaker
%binding affinity give larger area per molecule 
%increasing counterion binding affinity reduces the area per molecule
%when comparing between sodium and potassium in 
%or between
%ff99 and jc ions in Lipid 17 simulations.
In conclusion, the results are in line with the previous study suggesting that the
low area per molecule in PS lipid bilayers originate from the combination
of both counterion binding and hydrogen bonding network between lipid
headgroups \cite{petrache04}.
%, Berger and Lipid17/JC simulations.
%Slipids, Berger, MacRog, and Lipid17 exhibit a clear single maxima in the
%counterion density, but slightly at different distances from the bilayer
%center. Other models exhibit more complicated distribution, often including
%two local maxima in the density profile.
%CHARMM36, CHARMM36ua and Gromos-CKP models exhibit two local maxima in counterion
%density, while a single maxima is observed in the other models.

@@
